**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach

### 2020-09-15 manuscript/ecosphere_resubmit/manuscript.Rmd
edits done

---
title: "**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly mean anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. This analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale. Macro-scale (i.e., subcontinental) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale]. This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the nutrients [@collins2017lake]. On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern and midwestern US [@filstrup2014regional; @filstrup2018evidence]. Thus, it was surprising when little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader region of the northeastern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selction criteria on the LAGOSNE dataset as was applied to the URIWW data.  The LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013.

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term mean over a pre-determined reference period.  This allowed all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly means are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly mean anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

1. For each site, calculate the annual means, producing a single mean value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site means across years are assumed to be independent.
2. Calculate the long-term reference mean for each site. This results in a single long-term mean for each of the sites.
3. Calculate the anomaly for each annual mean at each site by subtracting the annual and reference means.
4. Summarize by calculating the mean anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

@@

*Linear regression on annual mean anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0). The slope of this line provides information on the mean yearly change of that parameter over the time period studied.

*Assessing regressions for trends:* Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence. For this assessment, we are defining a trend as positive or negative with the following; 1) Direction of trend is determined by the sign of the slope, 2) p-value is used to inform interpretation and very small p-values suggest a statistically relevant trend, 3) relative frequency of high and low years in the beginning and end of the time frame, and 4) the magnitude of the slope suggests an ecologically relevant change.

@@

During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a mean lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, mean TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).

For lakes and reservoirs in the larger region represented by the LAGOSNE States,  mean TN was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, mean TP was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, mean TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and mean chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Mean), lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).

@@

Mean annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests a statistical trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with mean temperature greater than the long-term mean are occurring in recent years (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (mean anomaly) in Rhode Island lakes and reservoirs. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of 0.3 μg/L is much greater than our threshold of 0.035 μg/L and the above-average years have all occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (mean anomaly).  Panel A. Yearly mean chlorophyll a anomalies from the URI Watershed Watch data.  Panel B. Yearly mean chlorophyll a anomalies from the LAGOSNE dataset. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means.", display = "cite")`A.).

```{r no_early_tn, echo=FALSE}
ww_tn_rl_noearly <- lm(mn_value~year, data = ww_tn_df[c(-1,-2),]) %>%
  tidy() %>%
  slice(2) %>%
  select(slope = estimate, p.value)
```

```{r no_early_tp, echo=FALSE}
ww_tp_rl_noearly <- lm(mn_value~year, data = ww_tp_df[c(-1,-2,-3),]) %>%
  tidy() %>%
  slice(2) %>%
  select(slope = estimate, p.value)
```

Mean annual trends for nutrients were not as clear. For total nitrogen, the slope of the line is positive with some statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) and the years greater than the mean are somewhat more frequent in more recent years.  However, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (mean anomaly). Panel A. Yearly mean TN anomalies from the URI Watershed Watch dataset.  Panel B. Yearly mean TN anomalies from the LAGOSNE dataset. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means. Grey indicates yearly site specifica anomalies that were, on average equal to the long-term means.  Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).  We interpret this as a weak positive trend in TN, but with little ecological impact. Total phosphorus shows a positive trend over time in the yearly anomalies but with somewhat weak statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the mean are more common in recent years (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (mean anomaly).  Panel A. Yearly mean TP anomalies from the URI Watershed Watch dataset.  Panel B. Yearly mean TP anomalies from the LAGOSNE dataset. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means. Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.). Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is greater than our defined threshold of 0.1 μg/L and suggests a possible trophic state shift in less than a century.  The pattern for the TN:TP ratio has a negative slope but no statistical support for a meaningful trend (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (mean anomaly).  Panel A. Yearly mean TN:TP ratio anomalies from the URI Watershed Watch dataset.  Panel B. Yearly mean TN:TP ratio anomalies from the LAGOSNE dataset. Points are means of site-specific anomalies and ranges are standard deviations of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term means. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term means. Missing years had insufficient data to calculate mean and standard deviation and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.).  Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

The analysis approach we used here, site-specific anomalies, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g., @Oliver_2017], we chose mean site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.  For instance, reporting the changes in anomalies allows us to look at changes in the original units.  With our analysis, the slope of the regression line for temperature suggests a mean yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows a mean yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.

>>>>

---
title: "**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach"
author:
- ^*1*^ US Environmental Protection Agency, Office Of Research and Development, Atlantic
  Coastal Environmental Sciences Division, Narragansett, RI 02882
- ^*2*^ University of Rhode Island, Department of Natural Resources Science, Kingston,
  RI 02881
- ^*3*^ ORISE, Narragansett, RI 02882
- '**Authors:** Hollister. J. W. ^\*^ ^*1*^, Kellogg, D. Q. ^*2*^, Kreakie, B. J. ^*1*^, Shivers, S. ^*3*^, Milstead, W. B. ^*1*^, Herron, E. ^*2*^, Green, L. ^*2*^, Gold, A. ^*2*^'
- '**Corresponding author: hollister.jeff@epa.gov**'
bibliography: manuscript.bib
output:
  word_document:
    reference_docx: manuscript_template.docx
editor_options:
  chunk_output_type: console
csl: ecology.csl
abstract: |
  | **Abstract:** Addressing anthropogenic impacts on aquatic ecosystems is a focus of lake management. Controlling phosphorus and nitrogen can mitigate these impacts, but determining management effectiveness requires long-term datasets. Recent analysis of the LAke multi-scaled GeOSpatial and temporal database for the Northeast (LAGOSNE) United States found stable water quality in the northeastern and midwestern United States, however, sub-regional trends may be obscured. We analyze a sub-regional (i.e., 3000 km^2^) trend with the University of Rhode Island’s Watershed Watch Volunteer Monitoring Program (URIWW) dataset. URIWW has collected water quality data on Rhode Island lakes and reservoirs for over 25 years. The LAGOSNE and URIWW datasets allow for comparison of water quality trends at regional and sub-regional scales, respectively. We assess regional (LAGOSNE) and state (URIWW) trends with yearly median anomalies calculated on a per-station basis. Sub-regionally, temperature and chlorophyll *a* increased from 1993 to 2016. Total nitrogen shows a weak increase driven by low years in the early 1990s. Total phosphorus and the nitrogen:phosphorus ratio (N:P) were stable. At the regional scale, the LAGOSNE dataset shows similar trends to prior studies of the LAGOSNE with chlorophyll *a*, total nitrogen, total phosphorus, and N:P all stable over time. In short, algal biomass, as measured by chlorophyll *a* in Rhode Island lakes and reservoirs is increasing, despite stability in total nitrogen, total phosphorus, and the nitrogen to phosphorus ratio. This analysis suggests an association between lake temperature and primary production. Additionally, we demonstrate both the value of long-term monitoring programs, like URIWW, for identifying trends in environmental condition, and the utility of site-specific anomalies for analyzing for long-term water quality trends. 
  | **Keywords:** Volunteer Monitoring, Citizen Science, Nutrients, Lakes, Trend Analysis, Long Term
---

@@

Previous studies using regional data from the northeastern and midwestern United States (US) have investigated spatial and temporal water quality trends and have shown differences based on scale. Macro-scale (i.e., ~ 1,800,000 km ^2^) drivers of water quality trends are complex and may vary temporally [@lottig2017macroscale]. This complexity can cause nutrient (N and P) trends to have different drivers than ratios of the nutrients [@collins2017lake]. On a regional scale, trends of N, P, and chlorophyll *a* differ as factors such as land use and climate vary between regions, particularly when comparing the northeastern (i.e. ~400,000 km^2^) and midwestern (i.e. ~1,000,000 km^2^) US regions [@filstrup2014regional; @filstrup2018evidence]. Furthermore, little change in nutrients and chlorophyll *a* was reported over a 25 year period for these regions [@Oliver_2017]. Given what is known about long-term trends in water quality within the broader regions of the northeastern and midwestern United States (US), we were curious if the lack of trends was also present in water quality at a sub-regional scale, using data on the 3,000 km^2^ area that encompasses a number of Rhode Island lakes and reservoirs.

@@

Prior studies have modeled water quality trends across a larger region of the northeastern US that included 17 states including Minnesota, Wisconsin, Iowa, Missouri, Illinois, Indiana, Michigan, Ohio, Pennsylvania, New York, New Jersey, Connecticut, Massachusetts, Rhode Island, Vermont, New Hampshire, and Maine [@soranno2015building; @Oliver_2017]. We repeated our analysis (see **Water Quality Trend Analysis** section) with the same dataset used by Oliver et al. [-@Oliver_2017], the LAGOSNE dataset [@soranno2015building; @stacheleklagosne2017; @sorannolagosne2017; @sorannolimno2019].  Temperature data were not available, thus we examined trends, using our analytical methods, for TN, TP, TN:TP, and chlorophyll *a* from the LAGOSNE dataset.  We used the same selection criteria on the LAGOSNE dataset as was applied to the URIWW data. This ensured that both datasets represented the same seasonal time frame.  Exact measurement depth for the LAGOSNE data were not available; however, the water quality data are reported to be either surface or epilimnion samples [@sorannolagosne2017]. At this time, the LAGOSNE data do not extend beyond 2013 thus sites needed to have a sampling event in 1993-2002 and in 2003-2013.

Lastly, to facilitate a comparison of the context of each scale, we use data, also from LAGOSNE, on landscape composition, maximum lake depth, and lake area [@sorannolagosne2017].  We use the 500 meter buffer landscape composition for three classes from the 2011 National Land Cover Dataset: Agriculuture, Forest, and Developed.  The maximum lake depth is from various sources and lake area is calculated from the National Hydropgraphy Dataset waterbody polygons.  Additional details are available in are detailed in Soranno et al. [-@sorannolagosne2017].

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variable we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from at <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

1. For each site, calculate the annual medians, producing a single median value for each site and year. This step prevents bias from pseudoreplication of multiple measurements of the same site in a given year [@hurlbert1984pseudoreplication].  The per site medians across years are assumed to be independent.
2. Calculate the long-term reference median for each site. This results in a single long-term median for each of the sites.
3. Calculate the anomaly for each annual median at each site by subtracting the annual and reference medians.
4. Summarize by calculating the median anomaly per year for the entire group of sites.  The resultant values are analyzed for a trend over time.

@@

*Linear regression on annual median anomalies:* Testing for a regression slope being different than zero can be used to test for monotonic trends in water quality data [@helsel2002statistical].  We used these standard procedures to test for positive or negative trends in lake temperature, chlorophyll *a*, TN, TP and TN:TP.  For each parameter, we fit a regression line to the anomalies as a function of year and tested the null hypothesis that no trend existed (e.g., $\beta$~1~ = 0). The slope of this line provides information on the average yearly change of that parameter over the time period studied.

*Assessing regressions for trends:* Traditionally, trends would be determined by assessing "significance" but recent guidelines suggest not using arbitrary p-value cut-offs to assesses significance [@wasserstein2016asa].  Our interpretation of the trends attempts to follow this advice and we assess trends with multiple lines of evidence. For this assessment, we are defining a trend as with the following: 1) p-values are used to determine general levels of statistical support, 2) relative frequency of high and low years in the beginning and end of the time frame suggest an increasing or decreasing patter, and 4) the magnitude of the slope is used to infer an ecologically relevant change.  Using all of this information we determine that a trend exists if it meets at least 2 of the 3 criteria, a weak trend exists if it meets one of the criteria and no trend exists if it meets none of the criteria.

@@

Lakes and reservoirs in Rhode Island tended to be smaller and shallower, on average than lakes included in the complete LAGOSNE region (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).  Also, LAGOSNE lakes tended to occur in landscapes (i.e. a 500 meter buffer) with a higher average percent agriculture, less developed land, and comparable forested area than lakes in Rhode Island (`r tab_num("context_summ", display = "cite", caption = "Average landscape context and lake morphometry summary statistics for lakes in URIWW and LAGOSNE.")`).

During the period of 1993 to 2016, Rhode Island lakes and reservoirs in our dataset had a median lake temperature of `r paste(ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Temperature")%>%pull(Units))`, median TN of `r paste(ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP of `r paste(ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio of `r paste(ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* of `r paste(ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), ww_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("ww_params", display = "cite", caption = "Summary statistics for URI Watershed Watch data from 1993 to 2016.")`).

For lakes and reservoirs in the larger region represented by the LAGOSNE States,  median TN was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Total Nitrogen")%>%pull(Units))`, median TP was `r paste(lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Total Phosphorus")%>%pull(Units))`, median TN:TP ratio was `r paste(lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "N:P")%>%pull(Units))`, and median chlorophyll *a* was `r paste(lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Median), lagos_param_ranges%>%filter(Parameter == "Chlorophyll")%>%pull(Units))` (`r tab_num("lagos_params", display = "cite", caption = "Summary statistics for LAGOSNE data from 1993 to 2016.")`).

On average, the Rhode Island lakes and reservoirs show lower concentrations of nutrients with more nitrogen relative to phosphorus than in the larger LAGOSNE region (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Chlorophyll concentrations also show lower average concentrations in Rhode Island (`r tab_num("ww_params", display = "cite")`, `r tab_num("lagos_params", display = "cite")`).  Furthermore, the distribution of chlorophyll based trophic state also shows a larger percentage of oligotrophic and mesotrophic lakes in Rhode Island and more eutrophic and hypereutrophic lakes in the larger LAGOSNE region [`r tab_num("ts_table", display = "cite", caption = "Percentage of lakes and reservoirs in different chlorophyll based trophic states.")`; @nurnberg1996trophic].

@@

Median annual temperature anomalies in lakes and reservoirs appears to be increasing as the slope is greater than the threshold of 0.02 we identified, the p-value suggests some statistical support for a positive trend (slope = `r signif(ww_temp_rl$slope,2)` , p = `r signif(ww_temp_rl$p.value, 2)`), and the majority of years with median temperature greater than the long-term median are occurring in the second half of the time period (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature (median anomaly) in Rhode Island lakes and reservoirs. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`, `r tab_num(name = "trend_summ", caption = "Summary of long term water quality trends in the URIWW and LAGOSENE data.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope = `r signif(ww_chla_rl$slope,2)` , p = `r signif(ww_chla_rl$p.value, 2)`).  The slope of `r signif(ww_chla_rl$slope,2)` μg/L is greater than our threshold of 0.035 μg/L and the above-average years have mostly occurred in the most recent years (`r fig_num(name = "chla_trend", caption = "Yearly trend over 20+ years of chlorphyll a (median anomaly).  Panel A. Yearly median chlorophyll a anomalies from the URI Watershed Watch data.  Panel B. Yearly median chlorophyll a anomalies from the LAGOSNE dataset. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

```{r no_early_tn, echo=FALSE}
ww_tn_rl_noearly <- lm(mdn_value~year, data = ww_tn_df[ww_tn_df$n > 3,]) %>%
  tidy() %>%
  slice(2) %>%
  select(slope = estimate, p.value)
```

```{r no_early_tp, echo=FALSE}
ww_tp_rl_noearly <- lm(mdn_value~year, data = ww_tp_df[c(-1,-2,-3),]) %>%
  tidy() %>%
  slice(2) %>%
  select(slope = estimate, p.value)
```

Median annual trends for nutrients were not as clear. For total nitrogen, the slope of the line is positive with weak statistical support (slope = `r signif(ww_tn_rl$slope,2)` , p = `r signif(ww_tn_rl$p.value, 2)`) but the years greater than the median are distributed evenly throughout time.  Also, a slope of `r signif(ww_tn_rl$slope,2)` μg/L per year is lower than our threshold value of 3.5 μg/L suggesting little support for a meaningful ecological change over time (`r fig_num(name = "tn_trend", caption = "Yearly trend over 20+ years of TN (median anomaly). Panel A. Yearly median TN anomalies from the URI Watershed Watch dataset.  Panel B. Yearly median TN anomalies from the LAGOSNE dataset. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians. Grey indicates yearly site specifica anomalies that were, on average equal to the long-term medians.  Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A.). However, 1998 only had three sites with available total nitrogen data and may not be representative.  This year also recorded the lowest median value.  Combined, 1998 may be skewing these results.  When this year is removed, the slope of the line is `r signif(ww_tn_rl_noearly$slope,2)` and the p-value is `r signif(ww_tn_rl_noearly$p.value, 2)`.  Thus, we interpret this as no trend in total nitrogen (`r tab_num(name = "trend_summ", display = "cite")`). Total phosphorus shows essentially no trend over time in the yearly anomalies with no statistical support (slope = `r signif(ww_tp_rl$slope,2)` , p = `r signif(ww_tp_rl$p.value, 2)`) and years that are over the median do not show any pattern (`r fig_num(name = "tp_trend", caption = "Yearly trend over 20+ years of TP (median anomaly).  Panel A. Yearly median TP anomalies from the URI Watershed Watch dataset.  Panel B. Yearly median TP anomalies from the LAGOSNE dataset. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians. Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`). Also a change of `r signif(ww_tp_rl$slope,2)` μg/L per year is much less than our defined threshold of 0.1 μg/L.  The pattern for the TN:TP ratio has a small slope and no statistical support for a meaningful trend (slope = `r signif(ww_np_rl$slope,2)`, p = `r signif(ww_np_rl$p.value,2)`) suggesting little evidence for a change in the concentrations of TN relative to the concentrations of TP (`r fig_num(name = "np_trend", caption = "Yearly trend over 20+ years of the TN:TP ratio (median anomaly).  Panel A. Yearly median TN:TP ratio anomalies from the URI Watershed Watch dataset.  Panel B. Yearly median TN:TP ratio anomalies from the LAGOSNE dataset. Points are medians of site-specific anomalies and ranges are the 25th and 75th percentiles of site-specific anomalies. Blue indicates yearly site-specific anomalies that were, on average, below the site-specific long-term medians. Red indicates yearly site-specific anomalies that were, on average, above the site-specific long-term medians. Missing years had insufficient data to calculate median and the 25th and 75th percentiles and an asterisk indicates year with only 3 sites and error bars are the range of the data.", display = "cite")`A., `r tab_num(name = "trend_summ", display = "cite")`).

@@

The analysis approach we used here, site-specific anomalies, is not a novel method and does have a long history in the analysis of trends in climate [@hansen2006global; @hansen2010global; @jones1996calculating; @jones1999surface]. However, using it to examine water quality trends is a new application of the technique, as we could find little evidence of using it specifically for water quality trends.  We built on these methods and adapted them for use with long-term water quality trends.  While other methods are valid and robust [e.g., @Oliver_2017], we chose median site-specific anomalies as they can provide readily interpretable results, especially for communicating to general audiences.  For instance, reporting the changes in anomalies allows us to look at changes in the original units.  With our analysis, the slope of the regression line for temperature suggests an average yearly increase of `r signif(ww_temp_rl$slope, 2)` °C and the slope of the regression line for chlorophyll *a* shows an average yearly increase of `r signif(ww_chla_rl$slope, 2)` µg/l.
